DRAFT: Harmonized and trait aggregation paper

### 2020-10-04 Paper/Draft_harmonization_paper.tex
04.10: First res base simulation 04.10: First res base simulation

We suspect that the heterogeneity of information in freshwater invertebrate trait databases is likely a major reason for the lack of studies across biogeographic regions. To harmonize grouping features from different regions, first commonly accepted and unambiguous trait definitions are required \cite{schneider_towards_2019}. In the best case, grouping features would be classified into the same traits across databases or they could easily be harmonized using standardized terminology. However, a lack of standardized terminology of trait definitions and poor metadata quality in many trait databases is a common issue throughout the field of trait-based ecology \cite{baird_toward_2011, schneider_towards_2019}. Secondly, consistent coding of traits facilitates the compatibility of trait data from different databases. Traits can be binary (i.e. trait exists or not), continuous, or fuzzy coded variables. Fuzzy codes represent the affinity of an organism to express a certain trait. They are used to account for plasticity in traits and are usually converted to percentages. Continuous values are typically used for grouping features that can be measured, like body size. However, invertebrate trait databases are heterogeneous with regard to the coding they use for their traits \cite{culp_incorporating_2011} (Table \ref{tab:trait_databases_coding_differentiation}). Brown et al. \cite{brown_functional_2018} harmonized grouping features based on trait databases from Europe, North America, and New Zealand because in these trait databases identical grouping features are classified differently into traits. As the traits from North America were coded binary in contrast to the fuzzy coded traits from from Europe and New Zealand, the authors consulted experts to assign fuzzy coded traits to North American taxa or inferred them from the European trait database. Thus, it becomes apparent that using invertebrate trait data from several regions requires extensive data processing prior to the actual data analysis. A centralized database with standardized and unambiguous traits and a consistent coding of traits would minimize data processing effort.

@@

Aggregated trait affinities of the five trait aggregation methods (\textit{direct\_agg\textsubscript{median}}, \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{median}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_agg}) were compared to trait affinities assigned at family-level by experts, which were available for the Australian and North American database for a subset of grouping features and taxa. For the Australian database, we compared aggregated trait affinities with assigned trait affinities resolved at family-level for the grouping features feeding mode and size using data from Chessman et al. \cite{chessman_dissolved-oxygen_2018}. In Chessman et al. \cite{chessman_dissolved-oxygen_2018} feeding mode is classified similarly as in the harmonized Australian database except that the trait parasite is missing. We conducted the comparison for the 220 families available in Chessman et al. \cite{chessman_dissolved-oxygen_2018}. Considering each factor combination of family and trait (in total 8) this amounts to 1760 cases.

@@

% Stat. techniques developed -> cannot use full potential
% if data quality is poor, when synthesizing information is hard 
% In general, differences produced by the aggregation methods seem to arise by using mean or median, not driven by different weighting approaches (direct vs. stepwise vs. weighted)

@@

% Trait definitions 
\subfile{Trait_def_table/SI_trait_def_table2.tex}

@@

%%%%%%%%%%%%%%%%%%%%%%%% Examples of trait databases %%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % ? Also mention Tomanova? -> not used here
% \begin{table}[ht]
%   \centering
%     \caption{Examples of aquatic invertebrate trait databases established in the last decades. Differences in codings and trait definitions are outlined for these databases in Table \ref{tab:trait_databases_coding_differentiation} and Table S\ref{tab:trait_definitions}.}
%     \label{tab:overview_trait_dbs}
%     \begin{tabular}{lcccl}
%     \hline
%     Database & \specialcell{Nr. of grouping \\ features} & \specialcell{Main organism \\ groups} & \specialcell{Taxa and \\ resolution} & Citation \\ 
%     \hline
%     Freshwaterecology & 40 & \specialcell{Invertebrates, biased \\ towards aquatic insects} & Species-level & \cite{schmidt-kloiber_www.freshwaterecology.info_2015} \\ 
%     Tachet & 21 & \specialcell{Invertebrates, biased \\ towards aquatic insects} & \specialcell{Taxa on species, \\ genus, and family-level} & \cite{usseglio-polatera_biomonitoring_2000} \\ 
%     \specialcell{North America \\ (Vieira)} & 62 & \specialcell{Invertebrates, biased \\ towards aquatic insects} & \specialcell{Taxa on species, \\ genus, and family-level} & \cite{vieira_database_nodate} \\ 
%     \specialcell{North America \\ (Twardochleb)} & 11 & Aquatic insects & \specialcell{Taxa on species, \\ genus, and family-level} & \cite{vieira_database_nodate} & \textit{In prep.} \\ 
%     Australia & & \specialcell{Invertebrates, biased \\ towards aquatic insects} & \specialcell{Taxa on species, \\ genus, and family-level} & \cite{kefford_integrated_2020} \\
%     New Zealand & & & & \cite{Philips_and_Smith_NZ_DB_2018}\\
%     \hline
%     \end{tabular}
% \end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

We suspect that the heterogeneity of information in freshwater invertebrate trait databases is likely a major reason for the lack of studies across biogeographic regions. To harmonize grouping features from different regions, first commonly accepted and unambiguous trait definitions are required \cite{schneider_towards_2019}. In the best case, grouping features would be classified into the same traits across databases or they could easily be harmonized using standardized terminology. However, a lack of standardized terminology of trait definitions and poor metadata quality in many trait databases is a common issue throughout the field of trait-based ecology \cite{baird_toward_2011, schneider_towards_2019}. Secondly, consistent coding of traits facilitates the compatibility of trait data from different databases. Traits can be binary (i.e. trait exists or not), continuous, or fuzzy coded variables. Fuzzy codes represent the affinity of an organism to express a certain trait. They are used to account for plasticity in traits and are usually converted to percentages. Continuous values are typically used for grouping features that can be measured, like body size. However, invertebrate trait databases are heterogeneous with regard to the coding they use for their traits \cite{culp_incorporating_2011} (Table \ref{tab:trait_databases_coding_differentiation}). Brown et al. \cite{brown_functional_2018} harmonized grouping features based on trait databases from Europe, North America, and New Zealand because in these trait databases identical grouping features are classified differently into traits. As the traits from North America were coded binary in contrast to the fuzzy coded traits from Europe and New Zealand, the authors consulted experts to assign fuzzy coded traits to North American taxa or inferred them from the European trait database. Thus, it becomes apparent that using invertebrate trait data from several regions requires extensive data processing prior to the actual data analysis. A centralized database with standardized and unambiguous traits and a consistent coding of traits would minimize data processing effort.

@@

Aggregated trait affinities of the five trait aggregation methods (\textit{direct\_agg\textsubscript{median}}, \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{median}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_agg}) were compared to trait affinities assigned at family-level by experts, which were available for the Australian and North American database for a subset of grouping features and taxa. For the Australian database, we compared aggregated trait affinities with assigned trait affinities resolved at family-level for the grouping features feeding mode and size using data from Chessman \cite{chessman_dissolved-oxygen_2018}. In Chessman \cite{chessman_dissolved-oxygen_2018} feeding mode is classified similarly as in the harmonized Australian database except that the trait parasite is missing. We conducted the comparison for the 220 families available in Chessman \cite{chessman_dissolved-oxygen_2018}. Considering each factor combination of family and trait (in total 8) this amounts to 1760 cases.

@@

% Influence of aggregation methods on trait-environment relationships:
% 1) In general, differences produced by the aggregation methods seem to arise by using mean or median, 
% not driven by different weighting approaches (direct vs. stepwise vs. weighted) 
% -> link to taxonomic hierarchy, many families have only a few genera
% -> Similarly, many genera have only a few species
% -> linear relationship: the more genus per family, the more species (although there are exceptions -> e.g. Europe)
% 
% Differences for NOA greater due to coding?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Ben's Comment: -> But another issues is that these aggregation methods were being compared to traits assigned at the family level, which I understand was often done by expert opinion. So, these traits assigned at the family level may not be really “correct”! The difference between the trait aggregated to family and its designation at the family level could indicate: (a) the aggregation method is “no good”, (b) the trait assigned at the family level was “no good”, or (c) both a and b.

% Simulation result: 
% -  base simulation with similar amount of speices per genus and genera per family and varying trait values
% results can vary when variation in trait values increases 
% However, mostly the different aggregation methods result in similar trait affinities
% when differences occur, only for: 
% dir_median vs stepwise median
% direct mean vs stepwise median
% stepwise mean vs stepwise median
% direct mean vs direct median
% stepwise median vs weighed agg
% Trait variability: 0.3 and greater
% Max diff ~ 0.2

% - Results for varying taxonomic hierarchy:
% Which levels?
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% 2) Harmonization: Minor influence on RDA species scores
% RDA species scores 
%   -> Normalized eigenvectors
%   -> dependent on nr. of variables? Nr. of response variables not really influential 
% We used only a small-scale study -> What's different to large-scale studies?

% Further aggr. after harmonization: no difference to the harmonized dataset
% ?Logical because trait aggr. methods themselves showed only minor differences

% Complex Stat. techniques developed -> cannot use full potential

@@

@@
